### 2020-04-13 thesis/content/abstract.tex
started with ltag definition + various minor fixes
improved intro and conclusion

% !TEX root = ../main.tex
%
\pdfbookmark[0]{Abstract}{Abstract}
\chapter*{Abstract}%
\label{sec:abstract}
\vspace*{-10mm}

>>>>

%!TEX root = ../main.tex
%
\pdfbookmark[0]{Abstract}{Abstract}
\chapter*{Abstract}%
\label{sec:abstract}
\vspace*{-10mm}

This thesis describes the research field of \acl{gcr} from the perspective of the \ac{lta} problem.
It formally characterizes a selection of state-of-the-art \textit{graph kernels} and \acp{gnn} as instances of \ac{lta}.
Those characterizations are shown to be limited by the way in which they decompose graphs.
To overcome this limitation, an avenue for a more ``\acs{lta}-like'' \ac{gnn} is provided in form of so-called \textit{learned edge filters}.
To realize edge filters, the novel \textit{2-\acs*{wl}-\acs*{gnn}} model is proposed;
it is inspired by the two-dimensional \ac{wl} algorithm and proven to be strictly more expressive than existing \ac{gnn} approaches which are bounded by the more restrictive one-dimensional \ac{wl} algorithm.

### 2020-04-13 thesis/content/chapter-conclusion.tex
started with ltag definition + various minor fixes
improved intro and conclusion

% !TEX root = ../main.tex
% chktex-file 46
\chapter{Conclusion}%
\label{sec:conclusion}

@@

>>>>

%!TEX root = ../main.tex
% chktex-file 46
\chapter{Conclusion}%
\label{sec:conclusion}

To conclude the thesis, we now look back on the three research questions described in \cref{sec:intro:questions} and summarize the answers we gave to them in the previous chapters.
Afterwards, a brief overview of future research directions based on our findings will be given.

@@

\paragraph{\circled{1}\; What constitutes an \ac{lta} method?}
We began with a general definition of \ac{lta} in \cref{sec:ltag:definition}.
There we proposed that its defining characteristic should be the \textit{localized explainability} of its predictions.
This characteristic was formalized via the notion of \textit{\ac{lta} formulations} (see \cfullref{defn:ltag:formulation}) which requires that a model is expressible in terms of a decomposition function $\psi: \mathcal{G} \to \mathcal{P}(\mathcal{G})$, a local evaluation function $f: \mathcal{G} \to \mathcal{Y} \times \mathcal{R}$ and a weighted aggregation function $\mathcal{A}: {(\mathcal{Y} \times \mathcal{R}_{\geq 0})}^* \to \mathcal{Y}$.
An ideal \ac{lta} method has such a formulation with a decomposition function $\psi$ that splits graphs into ``meaningful'' constituents in some domain-specific sense of the word.
Since this ideal notion of \ac{lta} is generally quite fuzzy, we only distinguished between \acs{lta}-like and non-\acs{lta} methods in this thesis;
a method was called non-\acs{lta} if it uses a trivial decomposition function that just splits a graph $G$ into the single ``constituent'' $G$.

\paragraph{\circled{2}\; How do existing \ac{gcr} methods relate to \ac{lta}?}
In \cref{sec:ltag:formulation:svm,sec:ltag:formulation:gcnn} we used our definition of \ac{lta} to check which of the existing \ac{gcr} approaches are compatible with it.
For the case of an \ac{svm} using a graph kernel/embedding we found that it is an \acs{lta}-like method if the kernel is a so-called nontrivial \acf{sce} (see \cfullref{defn:ltag:substruct-embedding,thm:ltag:svm-ltag-formulation}).
This \ac{sce} condition is satisfied by fingerprint embeddings, the \ac{wl} subtree kernel and the 2-LWL kernel which makes them \acs{lta}-like.
However, \texttt{graph2vec} embeddings, the \ac{wl} shortest-path kernel and the 2-GWL kernel were found to be trivial or only partly nontrivial \acp{sce}, i.e.\ they are non-\acs{lta} methods.
After considering those embedding approaches we looked at \ac{gcnn} and showed that they also have an \ac{lta} formulation under certain conditions (see \cfullref{thm:ltag:gcnn-ltag-formulation}).
More specifically, we saw that the constituents used by a \ac{gcnn} are spanned by the \ac{bfs} subtrees of its input graph.

\paragraph{\circled{3}\; What are limitations of existing graph \ac{lta} methods and how can they be overcome?}
\Cref{sec:ltd:edge-filter} described that the subtree constituents are their primary limitation of \acp{gcnn} and that more flexible decompositions can be learned via an edge filtering strategy.
To realize edge filtering, we proposed that informative edge feature vectors could be used as the input to a filtering classifier.
To produce such feature vectors we first looked at 2-\acsp{gnn} and found that they have various theoretical limitations, i.e.\ the inability to distinguish regular graphs and to detect cycles (see \cfullref{prop:ltd:2gnn-regular-limit,prop:ltd:2gnn-cycle-limit}).
We therefore proposed the 2-\acs{wl}-\acs{gnn} which does not have those limitations (see \cfullref{cor:ltd:wl2-gnn-regular}).

\paragraph{Evaluation results}
For the evaluation of our results we considered two aspects:
Firstly, we looked at how 2-\acs{wl}-\acsp{gnn} compare to other \acp{gnn}.
Secondly, we evaluated how \acs{lta}-like methods compare to non-\acs{lta} methods.
Regarding the first aspect, we showed that the theoretical advantages of 2-\acs{wl}-\acsp{gnn} are clearly observable on the synthetic triangle detection dataset while on the evaluated real-world datasets we got results which are generally comparable with the best state-of-the-art approaches but not significantly better.
Regarding the second aspect, we observed no general advantage or disadvantage of \acs{lta}-like methods.
While the \acs{lta}-like configurations of 2-\acs{wl}-\acsp{gnn} generally performed worse than their non-\ac{lta} counterparts, the \acs{lta}-like \ac{wl} subtree kernel generally performed quite well.
This shows that \ac{lta} is in principle suitable for graph classification tasks if the right decomposition, evaluation and aggregation functions are chosen.

### 2020-04-13 thesis/content/chapter-evaluation.tex
improved intro and conclusion

In \cref{sec:ltag} the relation between \ac{lta} and existing \ac{gcr} approaches was formally analyzed.
There we saw that the \ac{lta} formulations of existing approaches mostly use static decomposition functions, e.g.\ \ac{bfs} subtree decompositions.
Motivated by the idea of dynamically learning decompositions via edge filters, we then proposed the novel 2-\acs{wl}-\acs{gnn} in \cref{sec:ltd}.
The ides presented in both chapters will now be empirically evaluated.
Do to so we differentiate between two mostly independent evaluation aspects:
\begin{enumerate}[label={\textbf{\arabic*.}}]
	\item \textbf{Evaluation of 2-\acs{wl}-\acsp{gnn}:}
		Even though it was motivated by \ac{lta}, a 2-\acs{wl}-\acs{gnn} is not generally more ``\acs{lta}-like'' than other approaches.
		Nonetheless, due to the theoretical advantages described in \cref{sec:ltd:wl2gnn:properties} it is an interesting approach independently from its potential applications in \ac{lta} (see \cref{sec:ltd:edge-filter}).
		Thus the first aspect of our evaluation is to compare 2-\acs{wl}-\acsp{gnn} with the other previously described \ac{gcr} methods in a general non-\acs{lta} fashion, i.e.\ with an added \ac{mlp} after the pooling layer since this is how \acp{gnn} are typically evaluated in other works.
	\item \textbf{Evaluation of the \ac{lta} assumption:}
		We previously described that a given domain problem satisfies the \ac{lta} assumption if its solutions can be described by an \ac{lta} formulation (see \cfullref{defn:ltag:lta-assumption}).
		The inherent bias of an \acs{lta}-like model towards such \ac{lta} formulations could potentially increase its generalization performance compared to more general non-\acs{lta} models.
		Therefore the second aspect of our evaluation is to compare the performance of the previously described \acs{lta}-like methods with that of non-\acs{lta} approaches on datasets from multiple problem domains.
\end{enumerate}
This chapter will tackle those two aspects in four steps:
\begin{enumerate*}[label={\circled{\small\arabic*}}]
	\item We begin by describing the experimental setup used to obtain the evaluation results in \cref{sec:eval:setup}.
	\item We then present results on synthetically generated data in \cref{sec:eval:synthetic}.
	 	There we will illustrate the higher expressive power of 2-\acs{wl}-\acsp{gnn} when compared to other \ac{gcr} approaches which confirms the theoretical results from \cref{sec:ltd:wl2gnn:properties}.
	\item Then evaluation results on real-world datasets are described in \cref{sec:eval:real}.
		There we will see how 2-\acs{wl}-\acsp{gnn} compare to other \acp{gnn} in practice as well as how \acs{lta}-like models compare to non-\acs{lta} model.
	\item Finally, \cref{sec:eval:lta} looks at the synthetic and real-world results from an \ac{lta} perspective.
		There we will see how the predictive performance of an \acs{lta}-like model relates to the size and locality of its constituents.
\end{enumerate*}

@@

\begin{table}[ht]
	\caption[Mean test accuracies and standard deviations on real-world data.]{
		Mean test accuracies and standard deviations on real-world data.
		\hfill\textcolor{t_darkgreen}{*}\,\badgeboxinline[t_green]{\scriptsize\acs*{lta}-like}
	}\label{tbl:eval:real}
	\centering
	\csvreader[
		tabular={lrrrrr},
		separator=semicolon,
		table head={%
			& %
			\multicolumn{1}{c}{\hyperref[tbl:appendix:diff-nci]{\textbf{NCI1}}} &%
			\multicolumn{1}{c}{\hyperref[tbl:appendix:diff-proteins]{\textbf{PROTEINS}}} &%
			\multicolumn{1}{c}{\hyperref[tbl:appendix:diff-dd]{\textbf{D\&D}}} &%
			\multicolumn{1}{c}{\hyperref[tbl:appendix:diff-reddit]{\textbf{REDDIT}}} &%
			\multicolumn{1}{c}{\hyperref[tbl:appendix:diff-imdb]{\textbf{IMDB}}}%
			\\\toprule%
		},
		table foot=\bottomrule,
		late after line=\ifthenelse{\equal{\id}{9}}{\\\midrule}{\\},
		head to column names,
		filter={\equal{\isDefault}{1}}
	]{data/results.csv}{}{%
		% \ifthenelse{\equal{\id}{0}}{\multirow{5}{*}[0em]{\rotatebox[origin=c]{90}{\small\textsc{Kernel}}}}{}%
		% \ifthenelse{\equal{\id}{9}}{\multirow{8}{*}[0em]{\rotatebox[origin=c]{90}{\small\textsc{\ac{gnn}}}}}{} &%
		\textbf{\ifthenelse{\equal{\isLta}{1}}{\textcolor{t_darkgreen}{\model*}}{\model\ifthenelse{\equal{\model}{2-WL-GNN}}{\phantom{*}}{}}} {\footnotesize($\params$)} &%
		{\small\evalres{\nciBestTest}{\nciTestMean}{\nciTestStd}}&%
		{\small\evalres{\proteinsBestTest}{\proteinsTestMean}{\proteinsTestStd}}&%
		{\small\evalres{\ddBestTest}{\ddTestMean}{\ddTestStd}}&%
		{\small\evalres{\redditBestTest}{\redditTestMean}{\redditTestStd}}&%
		{\small\evalres{\imdbBestTest}{\imdbTestMean}{\imdbTestStd}}%
	}
\end{table}
Since the synthetic triangle detection problem was designed specifically to highlight the advantages of a higher dimensional \ac{wl} method such as 2-\ac{wl}-\acsp{gnn}, we now evaluate the approaches more fairly on five common graph classification benchmark datasets from two different real-world domains:
\begin{enumerate}[label={\textbf{\arabic*.}}]
	\item \textbf{Bioinformatics:}
		The NCI1~\cite{Shervashidze2011}, PROTEINS~\cite{Borgwardt2005a} and D\&D~\cite{Dobson2003} datasets all contain molecular graphs.
		For NCI1 the goal is to predict whether a molecule is effective against certain types of cancer.
		For PROTEINS and D\&D, a learner has to determine whether a given protein molecule is an enzyme.
	\item \textbf{Social networks:}
		The REDDIT and IMDB datasets~\cite{Yanardag2015} contain graphs representing the relations between users in online discussions and movie actors respectively.
		For REDDIT the type of a community has to be predicted based on a discussion thread; for IMDB the goal is to determine movie genres.
\end{enumerate}
A more detailed description of the evaluated datasets can be found in \cref{sec:appendix:ds-stats}.
\Cref{tbl:eval:real} shows our evaluation results\footnote{
	Note that, for the NCI1, PROTEINS and D\&D datasets we used different training/test splits than those originally used by \citet{Errica2020}.
	The reason for this is that the original splits were only published after our evaluations of those datasets had already been completed.
	For REDDIT and IMDB however, we used the same splits.
	Also note that some values are missing because of \acf{oom} errors;
	the resource use for all evaluations was limited to \SI{64}{\giga\byte} main memory and \SI{11}{\giga\byte} GPU memory.
	On average, each \ac{gnn} assessment (see \cref{algo:eval:assessment}) took about three days on an AMD 1800X 16-thread CPU and an Nvidia 1080~Ti GPU.% chktex 29
}.
For the evaluation of 2-\acs{wl}-\acsp{gnn}, different neighborhood radii were used for each dataset.
In the order of the columns in the table above, the results were obtained with the radii $r = 8$, $5$, $2$, $1$ and $4$ respectively.

>>>>

In \cref{sec:ltag} the relation between \ac{lta} and existing \ac{gcr} approaches was formally analyzed.
There we saw that the \ac{lta} formulations of existing approaches mostly use static decomposition functions, e.g.\ \ac{bfs} subtree decompositions.
Motivated by the idea of dynamically learning decompositions via edge filters, we then proposed the novel 2-\acs{wl}-\acs{gnn} in \cref{sec:ltd}.
The ideas presented in both chapters will now be empirically evaluated.
Do to so we differentiate between two mostly independent evaluation aspects:
\begin{enumerate}[label={\textbf{\arabic*.}}]
	\item \textbf{Evaluation of 2-\acs{wl}-\acsp{gnn}:}
		Even though it was motivated by \ac{lta}, a 2-\acs{wl}-\acs{gnn} is not generally more ``\acs{lta}-like'' than other approaches.
		Nonetheless, due to the theoretical advantages described in \cref{sec:ltd:wl2gnn:properties} it is an interesting approach independently from its potential applications in \ac{lta} (see \cref{sec:ltd:edge-filter}).
		Thus the first aspect of our evaluation is to compare 2-\acs{wl}-\acsp{gnn} with the other previously described \ac{gcr} methods in a general non-\acs{lta} fashion, i.e.\ with an added \ac{mlp} after the pooling layer since this is how \acp{gnn} are typically evaluated in other works.
	\item \textbf{Evaluation of the \ac{lta} assumption:}
		We previously described that a given domain problem satisfies the \ac{lta} assumption if its solutions can be described by an \ac{lta} formulation (see \cfullref{defn:ltag:lta-assumption}).
		The inherent bias of an \acs{lta}-like model towards such \ac{lta} formulations could potentially increase its generalization performance compared to more general non-\acs{lta} models.
		Therefore the second aspect of our evaluation is to compare the performance of the previously described \acs{lta}-like methods with that of non-\acs{lta} approaches on datasets from multiple problem domains.
\end{enumerate}
This chapter will tackle those two aspects in four steps:
\begin{enumerate*}[label={\circled{\small\arabic*}}]
	\item We begin by describing the experimental setup used to obtain the evaluation results in \cref{sec:eval:setup}.
	\item We then present results on synthetically generated data in \cref{sec:eval:synthetic}.
	 	There we will illustrate the higher expressive power of 2-\acs{wl}-\acsp{gnn} when compared to other \ac{gcr} approaches which confirms the theoretical results from \cref{sec:ltd:wl2gnn:properties}.
	\item Then evaluation results on real-world datasets are described in \cref{sec:eval:real}.
		There we will see how 2-\acs{wl}-\acsp{gnn} compare to other \acp{gnn} in practice as well as how \acs{lta}-like models compare to non-\acs{lta} models.
	\item Finally, \cref{sec:eval:lta} looks at the synthetic and real-world results from an \ac{lta} perspective.
		There we will see how the predictive performance of an \acs{lta}-like model relates to the size and locality of its constituents.
\end{enumerate*}

@@

\begin{table}[ht]
	\caption[Mean test accuracies and standard deviations on real-world data.]{
		Mean test accuracies and standard deviations on real-world data.
		\hfill\textcolor{t_darkgreen}{*}\,\badgeboxinline[t_green]{\scriptsize\acs*{lta}-like}
	}\label{tbl:eval:real}
	\centering
	\csvreader[
		tabular={lrrrrr},
		separator=semicolon,
		table head={%
			& %
			\multicolumn{1}{c}{\hyperref[tbl:appendix:diff-nci]{\textbf{NCI1}}} &%
			\multicolumn{1}{c}{\hyperref[tbl:appendix:diff-proteins]{\textbf{PROTEINS}}} &%
			\multicolumn{1}{c}{\hyperref[tbl:appendix:diff-dd]{\textbf{D\&D}}} &%
			\multicolumn{1}{c}{\hyperref[tbl:appendix:diff-reddit]{\textbf{REDDIT}}} &%
			\multicolumn{1}{c}{\hyperref[tbl:appendix:diff-imdb]{\textbf{IMDB}}}%
			\\\toprule%
		},
		table foot=\bottomrule,
		late after line=\ifthenelse{\equal{\id}{9}}{\\\midrule}{\\},
		head to column names,
		filter={\equal{\isDefault}{1}}
	]{data/results.csv}{}{%
		% \ifthenelse{\equal{\id}{0}}{\multirow{5}{*}[0em]{\rotatebox[origin=c]{90}{\small\textsc{Kernel}}}}{}%
		% \ifthenelse{\equal{\id}{9}}{\multirow{8}{*}[0em]{\rotatebox[origin=c]{90}{\small\textsc{\ac{gnn}}}}}{} &%
		\textbf{\ifthenelse{\equal{\isLta}{1}}{\textcolor{t_darkgreen}{\model*}}{\model\ifthenelse{\equal{\model}{2-WL-GNN}}{\phantom{*}}{}}} {\footnotesize($\params$)} &%
		{\small\evalres{\nciBestTest}{\nciTestMean}{\nciTestStd}}&%
		{\small\evalres{\proteinsBestTest}{\proteinsTestMean}{\proteinsTestStd}}&%
		{\small\evalres{\ddBestTest}{\ddTestMean}{\ddTestStd}}&%
		{\small\evalres{\redditBestTest}{\redditTestMean}{\redditTestStd}}&%
		{\small\evalres{\imdbBestTest}{\imdbTestMean}{\imdbTestStd}}%
	}
\end{table}
Since the synthetic triangle detection problem was designed specifically to highlight the advantages of a higher dimensional \ac{wl} method such as 2-\ac{wl}-\acsp{gnn}, we now evaluate the approaches more fairly on five common graph classification benchmark datasets from two different real-world domains:
\begin{enumerate}[label={\textbf{\arabic*.}}]
	\item \textbf{Bioinformatics:}
		The NCI1~\cite{Shervashidze2011}, PROTEINS~\cite{Borgwardt2005a} and D\&D~\cite{Dobson2003} datasets all contain molecular graphs.
		For NCI1 the goal is to predict whether a molecule is effective against certain types of cancer.
		For PROTEINS and D\&D, a learner has to determine whether a given protein molecule is an enzyme.
	\item \textbf{Social networks:}
		The REDDIT and IMDB datasets~\cite{Yanardag2015} contain graphs representing the relations between users in online discussions and movie actors respectively.
		For REDDIT the type of a community has to be predicted based on a discussion thread; for IMDB the goal is to determine movie genres.
\end{enumerate}
A more detailed description of the evaluated datasets can be found in \cref{sec:appendix:ds-stats}.
\Cref{tbl:eval:real} shows our evaluation results\footnote{
	Note that, for the NCI1, PROTEINS and D\&D datasets we used different training/test splits than those originally used by \citet{Errica2020}.
	The reason for this is that the original splits were only published after our evaluations of those datasets had already been completed.
	For REDDIT and IMDB however, we used the same splits.
	Also note that some values are missing because of \acf{oom} errors;
	the resource use for all evaluations was limited to \SI{64}{\giga\byte} main memory and \SI{11}{\giga\byte} GPU memory.
	On average, each \ac{gnn} assessment (as in \cref{algo:eval:assessment}) took about three days on an AMD 1800X 16-thread CPU and an Nvidia 1080~Ti GPU.% chktex 29
}.
For the evaluation of 2-\acs{wl}-\acsp{gnn}, different neighborhood radii were used for each dataset.
In the order of the columns in the table above, the results were obtained with the radii $r = 8$, $5$, $2$, $1$ and $4$ respectively.

### 2020-04-13 thesis/content/chapter-introduction.tex
started with ltag definition + various minor fixes
continued with spatial gcnn lta relation
reorganized thesis: wl2gnn gets its own chapter
includes the intro for the wl2gcnn chapter
+ various improvements
improved intro and conclusion

@@

The field of \ac{ml} on graph-structured data has applications in many domains due to the general expressive power of graphs.
The three most common types of graph~\ac{ml} problems are
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Link prediction:}
		A graph with an incomplete edge set is given and the missing edges have to be predicted.
		The suggestion of potential friends in a social network is a typical example for this.
	\item \textbf{Vertex classification \& regression:}
		Here a class or a score has to be predicted for each vertex of a graph.
		In social graphs this corresponds to the prediction of properties of individuals, e.g.\ personal preferences or gender.
		Another example is the prediction of the amount of traffic at the intersections of a street network.
	\item \textbf{Graph classification \& regression:}
		In this final problem type a single global class or continuous value has to be predicted for an input graph.
		The canonical example for this is the prediction of properties of molecule graphs, e.g.\ the toxicity or solubility of a chemical.
\end{enumerate}
In this thesis we will focus on the last problem type, \acf{gcr}.
An \ac{ml} method for this problem has to accept graphs of varying size and should be permutation invariant wrt.\ the order in which graph vertices are provided.
Those requirements are not met by the commonly used learners that only accept fixed-size feature vectors as their input, e.g.\ \acp{lm}, \acp{svm} or \acp{mlp}.

A \ac{gcr} method has to account for two central aspects of the problem:
\begin{enumerate*}
	\item Local structural analysis and
	\item global aggregation
\end{enumerate*}.
The first aspect is about the extraction of relevant features of substructures of the input graph.
The latter is about the way in which the local features are combined into a final class or regression value.
The existing \ac{gcr} methods are mostly motivated by local structural graph analysis.
The aspect of global aggregation on the other hand is less emphasized by those methods.

There is however a separate branch of research that specifically looks at the problem of learning aggregation functions, called \ac{lta}.
Current \ac{lta} approaches explicitly learn an aggregation functions for sets which can be interpreted as graphs without edges.
The motivation for this thesis is to generalize \ac{lta} from sets to arbitrary graphs.
The overall goal is to combine the aggregation learning perspective with existing \ac{gcr} methods.

\section{Goals}%
\label{sec:intro:goals}

To extend \ac{lta} to graphs, three goals have to be achieved:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Formalization of \ac{lta}:}
		Before \ac{lta} can be extended, its essential characteristics have to be defined.
		Those characteristics should provide the terminology to formally capture the differences and similarities between \ac{lta} and existing \ac{gcr} methods.
	\item \textbf{Give an \ac{lta} interpretation of \ac{gcr} methods:}
		Using the \ac{lta} formalization, representative \ac{gcr} approaches should be restated as \ac{lta} instances.
		Currently there is no comprehensive formulation of the relation between both fields of research;
		this is addressed by the the second goal.
	\item \textbf{Define an \ac{lta} method for graphs:}
		Using the \ac{lta} perspective on \ac{gcr}, hidden assumptions of the existing approaches should become clear and in which way they share the assumptions of \ac{lta}.
		The last goal is to use those insights to formulate an \ac{lta}-\ac{gcr} method that combines ideas from the existing approaches with the \ac{lta} assumptions.
\end{enumerate}

@@

\paragraph{\Cref{sec:related}: \nameref{sec:related}}

\paragraph{\Cref{sec:ltag}: \nameref{sec:ltag}}

\paragraph{\Cref{sec:eval}: \nameref{sec:eval}}

\paragraph{\Cref{sec:conclusion}: \nameref{sec:conclusion}}

>>>>

The field of \ac{ml} on graph-structured data has recently become an active topic of research.
One reason for this is the wide range of domains and problems that are expressible in terms of graphs.
Three very common types of such graph~\ac{ml} problems~\cite{Wu2019} are:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Link prediction:}
		A graph with an incomplete edge set is given and the missing edges have to be predicted.
		The suggestion of potential friends in a social network is a typical example for this.
	\item \textbf{Vertex classification \& regression:}
		Here a class or a score has to be predicted for each vertex of a graph.
		In social graphs this corresponds to the prediction of properties of individuals, e.g.\ personal preferences.
		Another example is the prediction of the amount of traffic at the intersections of a street network.
	\item \textbf{Graph classification \& regression:}
		In this final problem type, a single global class or continuous value has to be predicted for an input graph.
		The canonical example for this is the prediction of properties of molecular graphs, e.g.\ the toxicity or solubility of a chemical.
\end{enumerate}
In this thesis we will focus specifically on the last problem type, \ac{gcr}.
When comparing this problem with the more common problem of classification/regression on fixed-dimensional vector inputs $x \in \mathbb{R}^d$ we find two fundamental differences:
\begin{enumerate*}[label={\circled{\small\arabic*}}]
	\item The sizes of input graphs are generally not fixed; in \ac{gcr}, inputs can be arbitrarily small or large.
	\item Moreover, while two distinct vectors $x \neq x'$ typically represent distinct inputs for which different predictions $y \neq y'$ can be produced, graphs generally cannot be uniquely encoded because the order in which vertices are provided does not matter.
		A \ac{gcr} method must therefore be invariant w.r.t.\ vertex permutations, i.e.\ it must map two structurally identical graphs $G \simeq G'$ to the same prediction even if their encoding differs.
\end{enumerate*}

Due to those two differences, the \ac{gcr} problem cannot be directly solved by any of the common learners for vectorial inputs, e.g.\ \acp{lm} or \acp{mlp}.
Therefore, to tackle this problem, various graph-specific learners have been proposed over the recent years.
This thesis provides a novel perspective on the \ac{gcr} problem by combining the existing approaches with another, currently unrelated, field of research called \ac{lta}~\cite{Melnikov2016}\cite{Melnikov2019}.

@@

The existing \ac{gcr} approaches can be split into the family of so-called \textit{graph embeddings} and that of \acp{gnn}.
The embedding approach maps graphs to vectors in order to then apply a standard vector classification or regression method.
\Acp{gnn} on the other hand directly produce a prediction by using shared weights, similar to \acp*{cnn} on images~\cite{LeCun1998}.

Independent of those approaches, the family of \ac{lta} methods looks at the problem of aggregating variable-size \textit{compositions}, i.e.\ multisets, of so-called \textit{constituents}.
Like \ac{gcr} approaches, an \ac{lta} method has to be able to process inputs of varying size in a permutation invariant fashion.
Motivated by this similarity, this thesis extends the ideas of \ac{lta} to the domain of structured graph data.

Such an extension is interesting because \ac{lta} models produce composition predictions which are the direct result of local constituent predictions.
This means that \ac{lta} models provide insight into the influence of constituents on the final prediction, which makes \ac{lta} relevant from the perspective \ac{xai}~\cite{Gilpin2018}.
\Cref{fig:intro:intuition} illustrates how a graph-variant of \ac{lta} could explain the prediction of some graph property.
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.8\linewidth]{gfx/introduction/intuition.pdf}
	\caption{Intuition for how \ac{lta} could describe some graph property, e.g.\ the toxicity of a molecule, via a set of local constituent predictions.}\label{fig:intro:intuition}
\end{figure}

\section{Research Questions}%
\label{sec:intro:questions}

Based on the idea of extending \ac{lta} to graphs, we state the following three research questions which will be answered in this thesis:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Formalization of \ac{lta}:}
		The previous work on \ac{lta} provides only a narrow definition of \ac{lta} for unstructured multiset inputs.
		The question what its essential characteristics are, on a more general level, has not yet been considered.
		Therefore we will provide the terminology and definitions to formally capture what constitutes an \ac{lta} method as opposed to a non-\ac{lta} method.
	\item \textbf{An \ac{lta} interpretation of existing \ac{gcr} methods:}
		Using the generalized \ac{lta} formalization, a selection of state-of-the-art \ac{gcr} approaches will be checked for their compatibility with \ac{lta}.
		By doing so, we answer the question whether and how existing \ac{gcr} methods are able to produce predictions which are explainable by local constituent predictions.
	\item \textbf{Definition of a novel \acs{lta}-inspired \ac{gcr} method:}
		As a follow-up question to the previous one, we ask to which extent the existing \ac{gcr} methods are explainable and what their specific shortcomings are.
		Based on those shortcomings we will propose a novel \acs{lta}-inspired approach to overcome them.
\end{enumerate}

@@

\paragraph{\Cref{sec:related}: \nameref{sec:related}}
In order to answer our three research questions, we begin by with on overview of the previous work regarding \ac{lta} as well as \ac{gcr}.
We begin with a brief description of \ac{lta}.
Then the theoretical foundations of the existing \ac{gcr} approaches are introduced;
one particularly important concept there will be the so-called \acl*{wl} coloring.
Based on the theoretical foundations, we conclude the chapter by describing common graph embedding and \acl{gnn} methods.

\paragraph{\Cref{sec:ltag}: \nameref{sec:ltag}}
In this chapter we answer the first two of our research questions.
In the first step a formal definition of \ac{lta} is provided, which answers the question of what constitutes an \ac{lta} method.
This definition is then used to analyze all \ac{gcr} approaches described in \cref{sec:related} for their compatibility with \ac{lta}.

\paragraph{\Cref{sec:ltd}: \nameref{sec:ltd}}
Based on the \ac{lta} perspective on \ac{gcr} provided in \cref{sec:ltag}, we find that the existing approaches are all limited by their underlying solution to the so-called \ac*{ltd} problem.
To overcome this limitation, and thereby answer our third research question, we propose the general idea of edge-filtered graph convolutions.
To realize this general idea in a specific method, we then proceed by proposing a novel type of \ac*{gcnn}, the so-called \textit{2-\acs*{wl}-\acs{gnn}}, which can serve as the foundation for an edge-filtering-based solution to the \acs*{ltd} problem.
Apart from its uses in the context of \ac{lta}/\acs*{ltd}, we show that 2-\acs*{wl}-\acsp{gnn} have additional theoretical advantages over existing approaches;
this makes them interesting even from a non-\acs{lta} perspective.

\paragraph{\Cref{sec:eval}: \nameref{sec:eval}}
In this chapter we empirically evaluate how \ac{gcr} approaches that are compatible with our definition of \ac{lta} perform compared to non-\acs{lta} \ac{gcr} approaches.
Additionally, we compare the proposed 2-\acs*{wl}-\acs{gnn} with other state-of-the-art \acp{gnn} to verify whether its theoretical advantages over previous approaches can be observed in practice.

\paragraph{\Cref{sec:conclusion}: \nameref{sec:conclusion}}
Finally, the results of this thesis are summarized and a brief outline of promising directions for future research is given.

### 2020-04-13 thesis/content/chapter-related-work.tex
improved intro and conclusion

Before combining \ac{lta} and \ac{gcr} as described in \cref{sec:intro:goals}, we first give an overview of the state-of-the-art in both fields of research.
This is done in three steps:
\begin{enumerate}
	\item We begin with an overview of the existing \ac{lta} methods for unstructured inputs.
	\item Then we look at the domain of structured inputs.
		To solve the \ac{gcr} problem, relevant graphs characteristics have to be defined in order to determine the similarity and dissimilarity of graphs.
		We will look at three approaches for graph characterization: The graph isomorphism test, the so-called \acl{wl} coloring and lastly the notion of graph spectra.W
	\item Based on the described graph characterization approaches, an overview of current \ac{gcr} methods will then be given.
\end{enumerate}

@@

The last family of \ac{gcr} approaches we will look at is that of \acp{gnn}.
The idea to feed a graph into a \ac{nn} was first described by \citet{Gori2005}.
Since then many variants and extensions of that idea have been proposed~\cite{Wu2019}.
We will focus specifically on the so-called \acp{gcnn} which can be divided into two variants:
Spectral and spatial \acp{gcnn}.

>>>>

Before combining \ac{lta} and \ac{gcr}, as described in \cref{sec:intro:questions}, we first give an overview of the state-of-the-art in both fields of research.
This is done in three steps:
\begin{enumerate}
	\item We begin with an overview of the existing \ac{lta} methods for unstructured inputs.
	\item Then we look at the domain of structured inputs.
		To solve the \ac{gcr} problem, relevant graphs characteristics have to be defined in order to determine the similarity and dissimilarity of graphs.
		We will look at three approaches for graph characterization: The strict graph isomorphism test, the so-called \acl{wl} coloring and lastly the notion of graph spectra.
	\item Based on the described graph characterization approaches, an overview of current \ac{gcr} methods will then be given.
\end{enumerate}

@@

The last family of \ac{gcr} approaches we will look at is that of \acfp{gnn}.
The idea to feed a graph into a \ac{nn} was first described by \citet{Gori2005}.
Since then many variants and extensions of that idea have been proposed~\cite{Wu2019}.
We will focus specifically on the so-called \acp{gcnn} which can be divided into two variants:
Spectral and spatial \acp{gcnn}.
